---
title: "Computer Lab A"
subtitle: "Time Series Analysis -- 732A62"
author: "Jun Li"
date: '2019-09-21'
output: pdf_document
---

# Assignment 1. Computations with simulated data
## a) <!--example 1.9-->
```{r,eval=TRUE,echo=FALSE}
##Assignment 1 a)
set.seed(12345)
w=rnorm(100,0,1)
x=rep(NA,100)
x[1]=0
x[2]=0
for(i in 3:100) x[i]=x[i-2]+w[i]
vx=filter(x,filter=rep(0.2,5),sides=1)
y=rep(NA,100)
for(i in 1:100) y[i]=cos(2*pi*i/5)
vy=filter(y,filter=rep(0.2,5),sides=1)
print("Here is the comparation before and after filter for the first time series:")
par(mfrow=c(2,1))
plot.ts(x,main="The first series before filtering")
plot.ts(vx,main="The first series after filtering")
print("Here is the comparation before and after filter for the second time series:")
par(mfrow=c(2,1))
plot.ts(y,main="The second series before filtering")
plot.ts(vy,main="The second series after filtering")
```

## b) <!--example NA-->
```{r,eval=TRUE,echo=FALSE}
##Assignment 1 b)
rar=Mod(polyroot(c(1,-4,2,0,0,1)))>1
rma=Mod(polyroot(c(1,0,3,0,1,0,-4)))>1
rar
rma
```
The results above show if the roots are outside the unit circle. So the series is neither casual nor invertible.

## c) <!--example -->
After deduction on notes, it is shown that the series is stationay and the theoretical ACF is as follows: (deduction notes available if needed for aproval)

$$
\begin{aligned}
\rho(0)=1 \\
\rho(1)=-0.7089131 \\
\rho(h)=0.8478261*(\frac{-3}{4})^h  \\
\end{aligned}
$$

```{r,eval=TRUE,echo=FALSE}
##Assignment 1 c)
set.seed(54321)
w=rnorm(101,0,1)
x=rep(NA,100)
x[1]=0
for(i in 2:100) x[i]=-0.75*x[i-1]+w[i+1]-1/9*w[i-1]
print("Here is the sample ACF:")
acf(x,lag.max=10)
print("Here is the theoretical ACF:")
theo=rep(NA,11)
theo[1]=1
theo[2]=-0.7089131
for(i in 3:11) theo[i]=0.8478261*(-3/4)^(i-1)
plot(c(0:10),theo,type="h")
abline(h=0)
```

# Assignment 2. Visualization, detrending and residual analysis of Rhine data
## a) <!--Example 1.25-->
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
##Assignment 2 a)
data<-read.table(file="C:/Users/A550240/Desktop/LIU/TimeSeries/Rhine.csv",header=TRUE,sep=";",dec=",")
rh<-as.numeric(as.vector(data[,4]))
corr=round(acf(rh,12,plot=FALSE)$acf[-1],3) ## correlation between lag 0-12
par(mar=c(4,2,1,1))
plot.ts(rh,main="The series")
par(mfrow=c(6,2),mar=c(2,2,2,2))
for(i in 1:12) {ti=paste("rh-",i);plot(lag(rh,-i),rh,type="p",main=ti);legend("topleft",legend=corr[i])}
```
The series shows a seasonal trend as seen in the series plot, and has a highest concentration in december. There is a slightly decreasing trend in plot and variances seem to decrease as well. According to the autocorrelation results, the variables with lag -1,-2,-3,-10,-11,-12 have significant relation to the original series with higher correlation than 0.5.


## b)
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
##Assignment 2 b)
fit=lm(rh~time(rh),na.action=NULL)
re<-resid(fit)
par(mfrow=c(2,1))
plot(re,type="o",main="Residuals")
acf(re,24,main="ACF of residuals")
```
Residuals show a seasonal behaviour of time series, while ACF plot shows a positive correlation between months lagged by about 12 months and negative correlation between months lagged by about 6 months, which is seasonal as well.

## c) <!--Example 2.12 -->
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
##Assignment 2 c)
plot.ts(rh,main="Fitted by kernel smoother")
lines(ksmooth(time(rh),rh,"normal",bandwidth=20),lwd=2,col=4)
ksfit<-ksmooth(time(rh),rh,"normal",bandwidth=20)
re<-rh-ksfit$y
plot(re,main="Residuals")
acf(re,24,main="ACF of residuals")
```
It seems that residuals have a approximately constant mean 0 and ACF dependent only on lag. Therefore, it should be stationary.

## d) <!--Example 2.2+2.3+2.9-->
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
##Assignment 2 d)
du<-data[,2]
for(i in 1:12) assign(paste("du",i,sep=""),ifelse(du==i,1,0))
shu<-ts.intersect(rh,time(rh),du1,du2,du3,du4,du5,du6,du7,du8,du9,du10,du11,du12)
dufit<-lm(rh~time(rh)+du1+du2+du3+du4+du5+du6+du7+du8+du9+du10+du11+du12,data=shu)
re<-resid(dufit)
plot(re,main="Residuals")
acf(re,24,main="ACF of residuals")
```
It seems that residuals have a approximately constant mean 0. While the ACF plot shows a slight seasonal behaviour with a decreasing correlation as lag increases.

## e)

Here it is defined base mode as *rh~time(rh)*, then stepwise add *du1*,*du2* and so on. After every step, if the new AICc becomes lower, then keep the newly added variable, otherwise drop it and continue the next step. Here is the results as follows:
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
##Assignment 2 e)
newva="time(rh)"
dufit<-lm(rh~eval(parse(text=newva)),data=shu)
kn<-1  ## number of variables
aicold=log(sum(resid(dufit)^2)/168)+(168+kn)/(168-kn-2)
for(i in 1:12) 
  {temnewva=paste(newva,"+du",i,sep="")    ## added variables
   kn=kn+1
   dufit<-lm(rh~eval(parse(text=temnewva)),data=shu)
   temaic=log(sum(resid(dufit)^2)/168)+(168+kn)/(168-kn-2)
   if(temaic<aicold) {newva=temnewva;aicold=temnewva}
   else kn=kn-1}
print("Here comes the lowest AICc:")
aicold
print("Here comes the variables of best model:")
newva
```
The result shows that the best model is actually the base model, with AICc of  0.6166411.

# Assignment 3. Analysis of oil and gas time series
## a)
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
library(astsa)
library(ggplot2)
o<-as.data.frame(oil)
g<-as.data.frame(gas)  ##Preprocess as data frame
da<-cbind(o,g)
dimnames(da)[[2]]=c("oil","gas")
da$time<-c(1:545)
g<-ggplot(da,aes(x=time))+geom_line(aes(y=oil,colour="Oil"))+geom_line(aes(y=gas/2,colour="Gas"))+scale_y_continuous(sec.axis = sec_axis(~.*2, name = "Gas (cents/dollar)"))+ scale_colour_manual(values = c("blue", "red"))+labs(y = "Oil (dollars/barrel)",x = "Time",colour = "Variables")
g
```
Both series do not look like stationary since without transformation they seem to have changing mean and variances through time at least. And the two series probably have strong correlation, which in reality have similar properties as fossil fuel and are affected by the market in a similar way such as international demand and natural disasters.


## b)
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
o<-log(as.data.frame(oil))
g<-log(as.data.frame(gas))  ##Preprocess as data frame
da<-cbind(o,g)
dimnames(da)[[2]]=c("oil","gas")
da$time<-c(1:545)
g<-ggplot(da,aes(x=time))+geom_line(aes(y=oil,colour="Oil"))+geom_line(aes(y=gas,colour="Gas"))+scale_y_continuous(sec.axis = sec_axis(~., name = "Gas (cents/dollar)"))+ scale_colour_manual(values = c("blue", "red"))+labs(y = "Oil (dollars/barrel)",x = "Time",colour = "Variables")
g
```
The transformation converts two series with large unit difference into one unit demension, at the same time supresses the fluctuations so that two variables become comparable.

## c)
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
x_t<-diff(oil)
y_t<-diff(gas)
print("Here come the plots of detrended oil data:")
par(mfrow=c(2,1))
plot(x_t,main="Detrended oil")
acf(x_t,10,main="ACF of oil")
print("Here come the plots of detrended gas data:")
par(mfrow=c(2,1))
plot(y_t,main="Detrended gas")
acf(y_t,10,main="ACF of gas")
```
The ACF plots show that transformed oil data has stronger correlation with lag 1 and 8, while transformed gas data has stronger correlation with lag 3. In general, both series have approximate 0 mean. 

## d) <!--Example 2.13-->
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
par(mfrow=c(2,2))
for(i in -3:0) {plot(lag(x_t,i),y_t,main=paste("x_t,",i,spe=""));legend("topleft",legend=cor(y_t,lag(x_t,i)));lines(lowess(lag(x_t,i),y_t,f=10),lwd=2,col=4)}
```
It shows there is a significantly positive relation between series, and the trend seems stable through time. As shown in the figures, there several outliers in graphs.

## e)
```{r,eval=TRUE,echo=FALSE,warning=FALSE}
dux<-ifelse(x_t>0,1,0)
shu<-ts.intersect(y_t,dux,x_t,lag(x_t,-1),dframe=TRUE)
finfit<-lm(y_t~dux+x_t+lag(x_t,-1),data=shu)
summary(finfit)
re<-resid(finfit)
plot(re,main="Residuals")
acf(re,20,main="ACF of residuals")
```
The regression result shows that intercept and coefficients for *dux* and *x_t* are signicantly reliable since the t-value, the statistic index for hypothesis testing if the estimate is same as zero, is large enough. But not for $x_{t-1}$, since the NA indicates a colinearity with other variables. 
The residuals keep a approximately zero constant mean but an increasing variance under period 300-450, while ACF shows a week correlation with lagged series where lag 4 has a highest correlation anyway.



# Code Appendix

```{r code = readLines(knitr::purl("C:/Users/A550240/Desktop/LIU/TimeSeries/LabA.Rmd", documentation = 1)), echo = T, eval = F}
```