---
title: "Computer Lab C"
subtitle: "Time Series Analysis -- 732A62"
author: "Jun Li"
date: '2019-10-08'
output: pdf_document
---

# Assignment 1. Implementation of Kalman Filter
## a) <!--example -->
The state space model should be as follows:
\[z_t=z_{t-1}+e_t\]
\[x_t=z_t+v_t\]
\[e \sim N(0,1)\]
\[v \sim N(0,1)\]


## b) <!--example -->

```{r,eval=TRUE,echo=FALSE}
## b)
library(astsa)
# generate data
set.seed (1); num = 50
w = rnorm (num+1,0,1); v = rnorm (num,0,1)
mu = cumsum (w) # state : mu[0], mu[1] ,... , mu[50]
y = mu[-1] + v # obs: y[1] ,... , y[50]
# Moving average
vy<-filter(y,filter=rep(0.2,5),method="convolution",sides=1)
# filter and smooth ( Ksmooth0 does both )
ks = Ksmooth0(num , y, A=1, mu0=0, Sigma0=1, Phi=1, cQ=1, cR=1)
# start figure
Time = 1:num
plot (Time , mu[-1], main ='Filter ', ylim =c(-5,10))
lines (Time ,y,col=" green ")
lines (ks$xf)
#lines (ks$xf+2* sqrt (ks$Pf), lty =2, col=4)
#lines (ks$xf -2* sqrt (ks$Pf), lty =2, col=4)
lines (vy,lty =2, col=4)
```
The blue dashline is result from moving average, which comparing with Kalman filter is a biased fitting, and Kalman filter gives a better fit.

## c) <!--example -->

```{r,eval=TRUE,echo=FALSE}
## c)
# filter and smooth ( Ksmooth0 does both )
ks = Ksmooth0(num , y, A=1, mu0=0, Sigma0=1, Phi=1, cQ=1, cR=1)
# Modified filter
ksx = Ksmooth0(num , y, A=1, mu0=0, Sigma0=1, Phi=1, cQ=10, cR=0.1)
# start figure
par( mfrow =c(2,1));Time = 1:num
plot (Time , mu[-1], main ='Filter ', ylim =c(-5,10))
lines (Time ,y,col=" green ")
lines (ks$xf)
lines (ks$xf+2* sqrt (ks$Pf), lty =2, col=4)
lines (ks$xf -2* sqrt (ks$Pf), lty =2, col=4)
plot (Time , mu[-1], main ='Filter Modified', ylim =c(-5,10))
lines (Time ,y,col=" red ")
lines (ksx$xf)
lines (ksx$xf+2* sqrt (ks$Pf), lty =2, col=4)
lines (ksx$xf -2* sqrt (ks$Pf), lty =2, col=4)
```
The modified model with larger Q and small R presents a perfect fitting with original data.


## d) <!--example -->

```{r,eval=TRUE,echo=FALSE}
## d)
# filter and smooth ( Ksmooth0 does both )
ks = Ksmooth0(num , y, A=1, mu0=0, Sigma0=1, Phi=1, cQ=1, cR=1)
# Modified filter
ksx = Ksmooth0(num , y, A=1, mu0=0, Sigma0=1, Phi=1, cQ=0.1, cR=10)
# start figure
par( mfrow =c(2,1));Time = 1:num
plot (Time , mu[-1], main ='Filter ', ylim =c(-5,10))
lines (Time ,y,col=" green ")
lines (ks$xf)
lines (ks$xf+2* sqrt (ks$Pf), lty =2, col=4)
lines (ks$xf -2* sqrt (ks$Pf), lty =2, col=4)
plot (Time , mu[-1], main ='Filter Modified', ylim =c(-5,10))
lines (Time ,y,col=" red ")
lines (ksx$xf)
lines (ksx$xf+2* sqrt (ks$Pf), lty =2, col=4)
lines (ksx$xf -2* sqrt (ks$Pf), lty =2, col=4)
```
The modified model with small Q and larger R presents a heavily biased fitting with original data.

## e) <!--example -->

```{r,eval=TRUE,echo=FALSE}
## e)
myfilter<-setRefClass("myfilter",
  fields=list(num="numeric", y="vector", A="numeric", mu0="numeric", Sigma0="numeric", Phi="numeric", cQ="numeric", cR="numeric",xp="vector",xf="vector",xs="vector",Pp="vector",Pf="vector",Ps="vector"),
  
  methods=list(
    initialize=function(num="numeric", y="vector", A="numeric", mu0="numeric", Sigma0="numeric", Phi="numeric", cQ="numeric", cR="numeric",xp="vector",xf="vector",xs="vector",Pp="vector",Pf="vector",Ps="vector"){
      
      
      # Initialize
      ci<-num+1
      .self$xp<<-vector("numeric",ci);.self$xf<<-vector("numeric",ci);.self$xs<<-vector("numeric",ci)
      .self$Pp<<-vector("numeric",ci);.self$Pf<<-vector("numeric",ci);.self$Ps<<-vector("numeric",ci)
      .self$xp[1]<<-mu0;.self$xf[1]<<-mu0;.self$Pp[1]<<-Sigma0;.self$Pf[1]<<-Sigma0
      

      for(i in 2:ci){
      # Predict
      .self$xp[i]<<-Phi*.self$xf[i-1]
      .self$Pp[i]<<-Phi*.self$Pf[i-1]*t(Phi)+cQ
      # Filter
      k<-.self$Pp[i]*t(A)/(A*.self$Pp[i]*t(A)+cR) 
      .self$xf[i]<<-.self$xp[i]+k*(y[i-1]-A*.self$xp[i])
      .self$Pf[i]<<-(1-k*A)*.self$Pp[i]}
      
      
      # Smooth
      .self$xs[ci]<<-.self$xf[ci];.self$Ps[ci]<<-.self$Pf[ci] 
      for(i in num:2){
      g<-.self$Pf[i]*t(Phi)/.self$Pp[i+1]
      .self$xs[i]<<-.self$xf[i]+g*(.self$xs[i+1]-Phi*.self$xf[i])
      .self$Ps[i]<<-.self$Pf[i]+g*(.self$Ps[i+1]-.self$Pp[i+1])*t(g)}
    
        
      .self$xp<<-.self$xp[-1];.self$Pp<<-.self$Pp[-1]
      .self$xs<<-.self$xs[-1];.self$Ps<<-.self$Ps[-1]
      .self$xf<<-.self$xf[-1];.self$Pf<<-.self$Pf[-1]
      }
))


# filter and smooth ( Ksmooth0 does both )
ks<-myfilter$new(num , y, A=1, mu0=0, Sigma0=1, Phi=1, cQ=1, cR=1)

# start figure
par( mfrow =c(3,1)); Time = 1:num
plot (Time , mu[-1], main ='Predict ', ylim =c(-5,10))
lines (Time ,y,col=" green ")
lines (ks$xp)
lines (ks$xp+2* sqrt (ks$Pp), lty =2, col=4)
lines (ks$xp -2* sqrt (ks$Pp), lty =2, col=4)

plot (Time , mu[-1], main ='Filter ', ylim =c(-5,10))
lines (Time ,y,col=" green ")
lines (ks$xf)
lines (ks$xf+2* sqrt (ks$Pf), lty =2, col=4)
lines (ks$xf -2* sqrt (ks$Pf), lty =2, col=4)

plot (Time , mu[-1], main ='Smooth ', ylim =c(-5,10))
lines (Time ,y,col=" green ")
lines (ks$xs)
lines (ks$xs+2* sqrt (ks$Ps), lty =2, col=4)
lines (ks$xs -2* sqrt (ks$Ps), lty =2, col=4)
```
The hand-written function of Kalman Filter seems work well as it shows the same plots.


## f) <!--example -->
Kalman gain denotes the magnitude of taking correction of observations into consideration. As the covariance of $v_t$ is in the denominator, the larger $R_t$ is the unaccurate observation is, which leads to that the state estimates should rely less on observations, and the vice versa.




# Code Appendix

```{r code = readLines(knitr::purl("C:/Users/A550240/Desktop/LIU/TimeSeries/Task/LabC.Rmd", documentation = 1)), echo = T, eval = F}
```